Statement of Intent

### 2010-11-17 applications/nsf_fellowship/personal_statement.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

My father, grandfather, and great-grandfather all earned PhDs in physics.  I
grew up wanting to be a physicist, to win the Nobel Prize by discovering a
Unified Field Theory.  So of course, when I started college, I chose physics as
my major.  During my freshman year I was actively involved in the Society of
Physics Students, encouraging other students to study physics, helping to tutor
lower-level students, and even going to give physics and chemistry
demonstrations at a local middle school, hoping to excite the children to learn
about science.

@@

After my freshman year of college, I spent two years as a volunteer missionary
in Alabama, paying my way to serve the hispanic community in that state.  I
learned Spanish in order to communicate with them, and I came to understand
much better the circumstances that immigrants to our country live in.  I did a
significant amount of Spanish-English interpretation, and it was in that
capacity that I first started thinking about the potential for computer science
to improve the lives of ordinary people.  What if a computer could interpret
live speech from one language to another, accurately and effectively?  Such a
system would ease the burdens on so many immigrants who have much difficulty
communicating with other people for their daily needs, and not just those in
our country.

I came back from my experience in Alabama with a new-found desire to study
computer science.  I changed my major to computer science, with minors in
physics and mathematics, as those subjects were still interesting to me and
they seemed like relevant background.  I found a computer science professor who
was doing research that I found intriguing; I wanted to learn about machine
translation, which I knew was part of machine learning, so I started working in
the Applied Machine Learning Lab with Dr. Kevin Seppi.

My time working in the Applied Machine Learning lab has been the defining
experience of my undergraduate education.  I had very close interaction with
several professors and many graduate students who taught me what research was
like.  From those that I interacted with, I gained a love of learning new
things and exploring new ideas.  I also learned a great deal more than was
taught in my classes in order to understand the research that was going on in
the lab.  Several times in my classes the professor had me give part of a
lecture that dealt with things I was actively researching.

My undergraduate education would not have been nearly so fulfilling and
enjoyable without my experience working in a research lab.  My understanding of
the ideal college experience has been fundamentally changed through my
research.  If I ever end up as a professor, which I hope to be some day, one of
my goals will be to help all of the undergraduates in my program have the same
kind of experience that I had.  Working closely with professors and graduate
students and conducting independent research were wonderful opportunities that
every undergraduate student should be able to enjoy.

I also had a great internship experience as an undergraduate that helped to
shape my vision of what I wanted from my career.  I worked as an intern on the
TV Ads team at Google during the summer of 2009.  My intern host presented a
project to me that I didn't think was very interesting or very important, so I
proposed a more meaningful, and more difficult, project.  My host supported me
and gave me the help that I needed, and the project turned out to be very
successful.  What I learned from the experience was that while I enjoy writing
software, I find much more satisfication in solving interesting problems that
serve some useful purpose.

@@

My first semester as a masters student during the summer of 2010 gave me my
first real teaching opportunity.  While I have tutored other students, mostly
informally, throughout all of my college career, I had never done so in a
formal setting.  That semester, however, I was a teaching assistant for two
classes that Dr. Seppi taught: Introduction to Articial Intelligence, and
Bayesian Inference.  I had always enjoyed helping people understand concepts,
but I didn't think I would enjoy the teaching aspect of being a professor until
that experience.  I found that figuring out how to explain complicated concepts
simply and seeing the light of understanding in the students' faces gave me
great satisfaction.

My experience so far as a masters student has served to strengthen my desire to
get a PhD.  I love learning and solving new problems.  I envision my future
career as one of research, likely in the domain of natural language processing,
though possibly also in computer vision or other areas of machine learning.  I
have not yet decided between research in industry or academia.  I enjoy
teaching, but often the resources available in industry allow for more cutting
edge research; the massive amounts of data and extensive computing resources at
Google allow them to tackle problems that are impossible on smaller scales.
Perhaps my career will take me first to industry, ending up as a professor
later in life.  But what I most want from my career is to find new problems to
solve that will better society, then solve them.  Things like machine
translation and interpretation, or extracting knowledge from large collections
of text and making that knowledge accessible to people, are intriguing problems
that I would love to spend my life's work on.

>>>>

My father, grandfather, and great-grandfather all earned PhDs in physics.  I
grew up wanting to be a physicist, to win the Nobel Prize by discovering a
Unified Field Theory.  So of course, when I started college, I chose physics as
my major.  During my freshman year I was actively involved in the Society of
Physics Students, encouraging other students to study physics, helping to tutor
lower-level students, and even giving physics and chemistry demonstrations at a
local middle school, hoping to excite the children to learn about science.

@@

After my freshman year of college, I spent two years as a volunteer missionary
in Alabama, paying my way to serve the hispanic community in that state.  I
learned Spanish in order to communicate with them, and I came to understand
much better the circumstances that immigrants to our country live in.  I did a
significant amount of Spanish-English interpretation, and it was in that
capacity that I first thought about the potential for computer science to
improve the lives of ordinary people.  What if a computer could interpret live
speech from one language to another, accurately and effectively?  Such a system
would ease the burdens on so many immigrants who have great difficulty
communicating with other people for their daily needs, and not just those in
our country.

I came back from my experience in Alabama with a new-found desire to study
computer science.  I changed my major to computer science, with minors in
physics and mathematics, as those subjects were still interesting to me and
they are relevant background for a computer scientist.  I found a computer
science professor who was doing research that I found intriguing; I wanted to
learn about machine translation, which I knew was part of machine learning, so
I started working in the Applied Machine Learning Lab with Dr. Kevin Seppi.

My time working in the Applied Machine Learning lab has been the defining
experience of my undergraduate education.  I had very close interaction with
several professors and many graduate students who showed me how interesting
research can be.  From those that I had association with, I gained a love of
learning new things and exploring new ideas.  I also learned a great deal more
than was taught in my classes in order to understand the research that was
going on in the lab.  Several times in my classes the professor teaching the
class had me give part of a lecture that dealt with things I was actively
researching.

My undergraduate education would not have been nearly so fulfilling and
enjoyable without my experience working in a research lab.  My understanding of
the ideal college experience has been fundamentally changed through my
research.  As a professor, which I hope to be some day, one of my goals will be
to help all of the undergraduates in my program have the same kind of
experience that I had.  Working closely with professors and graduate students
and conducting independent research were wonderful opportunities that every
undergraduate student should be able to enjoy.

I also had a great internship experience as an undergraduate that helped to
shape my vision of what I wanted from my career.  I worked as an intern on the
TV Ads team at Google during the summer of 2009.  My intern host didn't have a
clear vision of the project she wanted me to do, so I proposed a meaningful,
difficult project that, if it worked, would have a large impact on the
company's business.  My host supported me and gave me the help that I needed,
and the project turned out to be very successful.  One of the main things I
learned from the experience was that while I enjoy writing software, I find
much more satisfication in solving interesting problems that actually improve
people's lives.

@@

My first semester as a masters student during the summer of 2010 gave me my
first real teaching opportunity.  While I have tutored other students, mostly
informally, throughout all of my college career, I had never done so in a
formal setting.  That semester, however, I was a teaching assistant for two
classes that Dr. Seppi taught: Introduction to Articial Intelligence, and
Bayesian Inference.  I had always enjoyed helping people understand concepts,
but I didn't realize how much I would like the teaching aspect of being a
professor until that experience.  I found that figuring out how to explain
complicated concepts simply and seeing the light of understanding in the
students' faces gave me great satisfaction.

My experience so far as a masters student has served to strengthen my desire to
get a PhD.  I love learning and solving new problems.  I envision my future
career as one of research, likely in the domain of natural language processing,
though possibly also in computer vision or other areas of machine learning.  I
have not yet decided between research in industry or academia.  I enjoy
teaching, but often the resources available in industry allow for more cutting
edge research; the massive amounts of data and extensive computing resources at
Google allow them to tackle problems that are impossible on smaller scales.
Perhaps my career will take me first to industry, then back to academia as a
professor later in life.  But what I most want from my career, wherever I work,
is to find new problems whose solutions will better society, then solve them.
Areas such as machine translation and interpretation, or extracting knowledge
from large collections of text and making that knowledge accessible to people,
are intriguing problems that I would love to spend my life researching.

### 2010-11-17 applications/nsf_fellowship/previous_research.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

\setlength{\textheight}{9in}

@@

This project started when a physics professor approached my advisor seeking
help solving a crystallography problem.  Having just begun working with my
advisor, I accepted the problem mostly in an auxiliary role, writing some code
and running experiments for Dr. Campbell, the physics professor.  We used
particle swarm optimization, a bio-inspired optimization algorithm similar to
evolutionary algorithms, to try to fit to data his model of what was happening
in a particular kind of crystal.  After spending a lot of time seeing the
output of the algorithm, I also helped to interpret the results.  I discovered
that we had too many free parameters in our model and helped Dr. Campbell to
make progress in solving his problem.  Because of our help in solving his
problem, Dr. Campbell has made what he considers a breakthrough in
understanding these crystals and plans to submit his work for publication soon.

@@

To solve this problem, my advisor proposed applying the idea of speculative
decomposition in the parallelization literature to particle swarm optimization
(PSO).  Instead of increasing the number of particles in the swarm, additional
processors could be used to evaluate possible next-iteration positions of each
particle in parallel.  I thought it was an interesting idea, and I made it my
honors thesis.  I took the idea and discovered how it could actually be done,
performing the evaluations needed and piecing the results back together from
messages sent between processors so that speculative evaulation could be done
in a distributed PSO environment with only a single round of message passing
per iteration.  After implementing the technique, I ran many experiments to
compare our methods to other means of parallelizing the algorithm.  Once I was
satisfied that my results were significant, I wrote, submitted, and
successfully defended my honors thesis.  I also modified the paper to fit in
the aims and page requirements of Parallel Problem Solving from Nature, a
well-respected parallel optimization conference.  I submitted my thesis
there and it was accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.

In implementing speculative evaluation, however, I noticed a number of ways the
method could potentially be improved.  Even though my honors thesis was
completed, I continued my investigation.  I found that a large number of the
speculative evaluations we performed were wasted, and that the algorithm could
be improved by replacing useless evaluations with better evaluations, such as
speculating several iterations ahead on paths that particles were likely to
take.  I significantly expanded the work of my honors thesis, preparing to
submit it to a journal on optimization.  After a rejection of a smaller version
from a conference, I got quite discouraged, wanting to give up and move on to
something new.  My advisor insisted I persevere, saying it was good preparation
for a PhD, and so I continued.  After running more experiments to strengthen
the paper, we submitted it to the journal Swarm Intelligence, where it is
currently under
review~\cite{gardner-2010-speculative-approach-to-parallelization-pso}.

While I have since move on to other interests, this project was my first
experience performing research and dealing with the submission process at
conferences.  All told, I worked on particle swarm optimization for about two
and a half years.  I learned a lot about what research is, the perseverance
required in getting a PhD, and how much I love exploring new topics and
extending the limits of what we can do, as well as helping researchers in other
areas to solve their problems.  This project set the stage for my masters
thesis and for my future work as a PhD student.

Along with the research I performed as an undergraduate research assistant,
there are two other research projects worth noting here.  One of them was my
internship at Google in the summer of 2009.  I was presented with the problem
of how to determine the effectiveness of TV advertisements given only a list of
at what times and on what channels the ads played, along with a list of phone
calls received in response to advertisements.  This was complicated because
often ads play at the same time on different channels, and they use the same
phone number.  I developed and implemented a probabilistic model that assigned
phone calls to ads that were played.  The model learned a response rate for
each channel the ads played on and helped advertisers know where to most
effectively spend their money.

The other research area that I have pursued is topic modeling, a method for
capturing the semantics of a document collection, and that is the area of my
masters thesis.  As an undergraduate, I implemented a topic model and used it
to analyze a novel dataset of religious discourses, which I presented at an
international conference held at
BYU~\cite{gardner-2010-general-conference-topics}.  As a graduate student, I
have been investigating how to use topic models to get specific, corpus-wide
information out of large collections of documents.  My initial endeavors have
involved building a tool to browse the output of a topic model in novel
ways~\cite{gardner-2010-topic-browser}.  As other departments at BYU have seen
the work that I and others have done with our topic browser, they have been
very interested in using our tool.  Three separate groups at BYU have asked to
use our browser to analyze data with purposes ranging from finding public
health information in Twitter feeds to finding problem areas in teaching from
free-form student ratings of classes and professors.  My future plans with
topic modeling are presented in the research proposal submitted with this
application.

\footnotesize
\bibliographystyle{plain}
\renewcommand\bibsection{\subsubsection*{Publications}}
\bibliography{../../../bib/aml/bib}

>>>>

\setlength{\textheight}{9in}
\setlength{\bibsep}{0in}

@@

This project started when a physics professor approached my advisor seeking
help solving a crystallography problem.  As a new student in the lab, I
accepted the problem mostly in an auxiliary role, writing some code and running
experiments for Dr. Campbell, the physics professor.  We used particle swarm
optimization (PSO), a bio-inspired optimization algorithm similar to
evolutionary algorithms, to try to fit to data his model of what was happening
in a particular kind of crystal.  After spending a lot of time seeing the
output of the algorithm, I also helped to interpret the results.  I discovered
that we had too many free parameters in our model and helped Dr.  Campbell to
make progress in solving his problem.  Because of our help in solving his
problem, Dr. Campbell has made what he considers a breakthrough in
understanding these crystals and plans to submit his work for publication soon.

@@

To solve this problem of diminishing returns, my advisor proposed applying the
idea of speculative decomposition in the parallelization literature to PSO.
Instead of increasing the number of particles in the swarm, additional
processors could be used to evaluate possible next-iteration positions of each
particle in parallel.  I thought it was an interesting idea, and I made it my
honors thesis.  I took the idea and discovered how it could actually be done,
performing the evaluations needed and piecing the results back together from
messages sent between processors so that speculative evaulation could be done
in a distributed PSO environment with only a single round of message passing
per iteration.  After implementing the technique, I ran many experiments to
compare our methods to other means of parallelizing the algorithm.  Once I was
satisfied that my results were significant, I wrote, submitted, and
successfully defended my honors thesis.  I also modified the paper to fit in
the aims and page requirements of Parallel Problem Solving from Nature, a
well-respected parallel optimization conference.  I submitted my thesis there
and it was accepted~\cite{gardner-2010-speculative-evaluation-in-pso}.

In implementing speculative evaluation, however, I noticed a number of ways the
method could potentially be improved.  Even though my honors thesis was
completed I continued my investigation, as I thought there was more that could
be done.  I found that a large number of the speculative evaluations we
performed were wasted, and that the algorithm could be improved by replacing
useless evaluations with better evaluations, such as speculating several
iterations ahead on paths that particles were likely to take.  I significantly
expanded the work of my honors thesis, preparing to submit it to a journal on
optimization.  After a rejection of a smaller version from a conference, I
became quite discouraged, wanting to give up and move on to something new.  My
advisor insisted I persevere, saying it was good preparation for a PhD, and so
I continued.  After running more experiments to strengthen the paper, we
submitted it to the journal Swarm Intelligence, where it is currently under
review~\cite{gardner-2010-speculative-approach-to-parallelization-pso}.

While I have since move on to other interests, this project was my first
experience performing research and dealing with the submission process at
conferences.  I learned a lot about what research is, the perseverance required
in getting a PhD, and how much I love exploring new topics and extending the
limits of what we can do, as well as helping researchers in other areas to
solve their problems.  This project set the stage for my masters thesis and for
my future work as a PhD student.

Along with the research I performed as an undergraduate research assistant, two
other projects are worth noting here.  The first is my internship at Google in
the summer of 2009.  I tackled the problem of how to determine the
effectiveness of TV advertisements given only a list of when and on what
channels the ads played, along with a list of phone calls received in response
to the ads.  This was complicated because often ads play at the same time on
different channels, and they use the same phone number.  I developed and
implemented a probabilistic model that assigned phone calls to ads that were
played.  The model learned a response rate for each channel the ads played on
and helped advertisers know where to most effectively spend their money.  Once
advertisers saw that following the recommendations of my model gave them better
response rates, they spent more on advertising with Google.

The other research area that I have pursued is topic modeling, a method for
capturing the semantics of a document collection, and that is the area of my
masters thesis.  As an undergraduate, I implemented a topic model and used it
to analyze a novel dataset of religious discourses, which I presented at an
international conference held at
BYU~\cite{gardner-2010-general-conference-topics}.  As a graduate student, I
have been investigating how to use topic models to get specific, corpus-wide
information out of large collections of documents.  My initial endeavors have
involved building a tool to browse the output of a topic model in new
ways~\cite{gardner-2010-topic-browser}.  As other departments at BYU have seen
the work that I and others have done with our topic browser, they have been
very interested in using our tool.  Three separate groups at BYU have asked to
use our browser to analyze data with purposes ranging from finding problem
areas in teaching from free-form student ratings of classes and professors to
searching for public health information in Twitter feeds.  My future plans with
topic modeling are presented in my research proposal.

\footnotesize
\bibliographystyle{plain}
\renewcommand\bibsection{\noindent \small\textbf{Publications}\vspace{-2mm}\footnotesize}
\bibliography{../../../bib/aml/bib}

### 2010-11-17 applications/nsf_fellowship/proposed_research.tex
Incorporated feedback from Sabrina, and I  am done with the previous research essay

An important aspect of the subset of the data that to be annotated will be that
it will consist of a set of documents that all have the same value for some
attribute, such as ratings for the same teacher, or reviews about the same
restaurant.  This will facilitate both the finding of consistent thematic
elements and more extensive kinds of validation, which will be explained later.

@@

\textbf{Broader Impacts.}  This work will quantitatively establish the
usefulness of topic models in extracting thematic elements from a collection of
documents.  It will also validate the automatic ranking of topics in response
to a few classes of specific user queries.  Such work will give confidence to
anyone using a topic model to analyze a corpus, including intelligence analysts
looking at classified documents, university administrators looking at student
ratings, and companies reading online reviews about their products.  This work
also establishes a method and a set of data for topic models to be compared to
each other, for one model may better capture that human-judged thematic
elements than another.

>>>>

An important aspect of the subset of the data that will be annotated is
that it will consist of a set of documents that all have the same value for
some attribute, such as ratings for the same teacher, or reviews about the same
restaurant.  This will facilitate both the finding of consistent thematic
elements and more extensive kinds of validation, which will be explained later.

@@

\textbf{Broader Impacts.}  This work will quantitatively establish the
usefulness of topic models in extracting thematic elements from a collection of
documents.  It will also validate the automatic ranking of topics in response
to a few classes of specific user queries.  Such work will give confidence to
anyone using a topic model to analyze a corpus, including intelligence analysts
looking at classified documents, university administrators looking at student
ratings, and companies reading online reviews about their products.  This work
also establishes a method and a set of data for topic models to be compared to
each other, for one model may better capture human-judged thematic elements
than another.
